---
title: 'Step 2.01: UF Outcome Creation'
author: "Mia Charifson"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/miacharifson/Library/CloudStorage/OneDrive-NYULangoneHealth/Charifson Dissertation/')

library(tidyverse)
library(ggplot2)
library(lubridate)
library(tableone)
library(arsenal)
library(readxl)
library(ggpubr)
library(corrplot)
library(data.table)
```

# 2.1. Derive outcome variables

Goals of this document: - Using algorithm to assign diagnosis confidence to each case - Identify patients for whom we need to manually review notes

```{r load}
pat_cohort <- read.csv('Data/Derived/pat_cohort_w_censoring.csv') 
uf_endo <- read.csv('Data/Epic/table3_dx_outcomes_new_20240304.csv')
uf_endo_filled_dates <-read.csv('Data/Epic/table3_dx_outcomes_date_filled_20240304.csv')
imaging_dx <- read.csv('Data/Epic/table3_dx_imaging.csv') 
imaging_orders <- read.csv('Data/Epic/table4_proc_imaging.csv') 
surgery_orders <- read.csv('Data/Epic/table4_proc_surgery.csv') 
surgery_records <- read.csv('Data/Epic/table4_surgery_surgical_dx.csv')
```

```{r}
pat_cohort %>%
  mutate(person_years = case_when(is.na(final_censor_date)~difftime('2023-08-01',index_enc_date,units='days'),
                                  TRUE~difftime(final_censor_date,index_enc_date,units='days')),
         person_years = as.numeric(person_years)/365.25) %>%
  summarise(total_person_time = sum(person_years))
```


```{r}
# Combine tables that need combining
uf_endo_cohort <- uf_endo %>%
  mutate(dx_date = strptime(dx_date, "%Y-%m-%d %H:%M:%S")) %>%
  ## remove those missing a dx date
  filter(!is.na(dx_date)) %>%
  ## add in filled dx_date where missing
  bind_rows(uf_endo_filled_dates %>% 
              filter(!is.na(dx_date)) %>%
              mutate(dx_date_filled = strptime(dx_date_filled, "%Y-%m-%d %H:%M:%S"),
                     dx_date = strptime(dx_date, "%Y-%m-%d %H:%M:%S")) %>%
              mutate(dx_date = case_when(is.na(dx_date)~dx_date_filled, TRUE~dx_date),
                     pat_enc_csn_id = as.character(pat_enc_csn_id))) %>%
  # fill in missing index_enc_date
  inner_join(pat_cohort %>% select(pat_id, index_enc_date), by = 'pat_id')
  
## combine all imaging
imaging_all <- imaging_dx %>%
  mutate(imaging_date = strptime(dx_date, "%Y-%m-%d %H:%M:%S")) %>%
  #mutate(pat_enc_csn_id = as.numeric(pat_enc_csn_id)) %>%
  mutate(event_name = dx_name,
         event_type = paste0('Imaging-', dx_type)) %>%
  select(pat_id, imaging_date, event_name, event_type) %>%
  bind_rows(imaging_orders %>% 
              mutate(imaging_date = as.Date(ordering_date)) %>%
              mutate(event_name = procedure_name,
                     event_type = 'Imaging Order') %>%
              select(pat_id, imaging_date, event_name, event_type)) %>%
  filter(pat_id %in% pat_cohort$pat_id)

## combine all surgery
surgery_all <- surgery_records %>%
  mutate(surgery_date = strptime(surgery_date, "%Y-%m-%d %H:%M:%S")) %>%
  full_join(surgery_orders %>% mutate(surgery_date = as.Date(ordering_date)), 
            by=c('pat_id', 'pat_enc_csn_id', 'surgery_date', 'procedure_name',
                                 'procedure_code_type', 'procedure_code')) %>%
  mutate(type = ifelse(!is.na(performed), 'Surgery record', 'Surgery order')) %>%
  select(-ordering_date) %>%
  filter(pat_id %in% pat_cohort$pat_id)
```

```{r, eval=FALSE}
# there should be no UF/Endo diagnoses prior to their index enc
length(which(uf_endo_cohort$dx_date[uf_endo_cohort$icd10 != 'N80.03'] < uf_endo_cohort$index_enc_date[uf_endo_cohort$icd10 != 'N80.03']))

# look at these patients
#uf_endo_cohort %>% filter(icd10 != 'N80.03' & dx_date <= index_enc_date) %>% View()
```

```{r, eval=FALSE}
## check how many of instances of the history code for UF is
uf_endo_cohort %>% filter(grepl('Z86', icd10)) ## will account for this later on
```

## Uterine fibroids

```{r unique_dates}
uf_other_codes <- c('D21.9', 'O34.10', 'O34.11', 'O34.12', 'O34.13', 'IMO0001', 'N93.9', 'D28.1', 'O46.8X1')
uf_names <- 'fibroid|leiomyoma|uterine myoma|uterine fibromyoma'

uf_dx_unique <- uf_endo_cohort %>% 
  filter(!is.na(dx_date)) %>%
  filter(grepl('D25', icd10) | (icd10 %in% uf_other_codes & grepl(uf_names, tolower(dx_name)))) %>%
  rename(date=dx_date) %>%
  select(-referral_id, -icd10, -dx_date_filled, -pat_enc_csn_id) %>%
  group_by(pat_id, date) %>%
  summarise(event_name = paste0(unique(dx_name), collapse=', '),
            event_type = paste0(unique(dx_type), collapse=', '),
            .groups='drop')

imaging_unique <- imaging_all %>% 
  mutate(date=as.Date(imaging_date)) %>%
  group_by(pat_id, date)  %>%
  summarise(event_name = paste0(unique(event_name), collapse=', '),
            event_type = paste0(unique(event_type), collapse=', '),
            .groups='drop')

surgery_unique <- surgery_all %>% 
  mutate(date = as.Date(surgery_date)) %>%
  group_by(pat_id, date) %>%
  summarise(event_name = paste0(unique(procedure_name), collapse=', '),
            event_type = paste0(unique(type), collapse=', '),
            .groups='drop')
```

```{r}
uf_all_long <- bind_rows(uf_dx_unique, imaging_unique, surgery_unique) %>%
  distinct() %>%
  arrange(pat_id, date)

n_uf_raw <- length(unique(uf_dx_unique$pat_id))
```

There are `r n_uf_raw` patients who have any UF Dx code added to their EHR at some point during follow-up.

The next steps are to determine how many of those diagnoses we feel are true positives as opposed to potential false positives. 

There are several "hints" in the EHR which might encourage confidence that a UF Dx is a true positive. These hints are:

- Having a diagnostic procedure
- Confirmation of UF Dx following a diagnostic procedure
- Any UF Dx after a diagnostic procedure
- Multiple UF Dx codes (not on the same day)
- Referral diagnosis
- UF symptoms
- Having only medical history diagnoses (negative hint for prevalent case)

I will look for these "hints" for each patient with a UF Dx and then based on some combination of those hints determine if we feel the Dx is a `confident positive` or `potential false positive`. 

I will also look for these "hints" for patients without a UF Dx (such as a diagnostic procedure or UF symptoms) to try to identify `confident negatives` and `potential false negatives`. 

`Confident positive` is defined as a UF diagnosis which has some combination of the following hints (in order of confidence):

1. Confirmation of UF Dx following a diagnostic procedure (same day as procedure = most confident)
2. Having a diagnostic procedure & any UF Dx after a diagnostic procedure
3. Multiple UF Dx codes (not on the same day)
4. Referral diagnosis

`Potential false positive` is defined as a patient with only one Dx code for UF in their EHR during follow-up and no diagnostic procedure and it was not a referral diagnosis.

### First diagnosis information

```{r uf_first_dx}
uf_first_dx <- uf_endo_cohort %>% 
  filter(!is.na(dx_date)) %>%
  filter(grepl('D25', icd10) | (icd10 %in% uf_other_codes & grepl(uf_names, tolower(dx_name)))) %>%
  group_by(pat_id) %>%
  filter(dx_date==min(dx_date)) %>%
  summarise(first_dx_date = mean(dx_date),
            first_dx_type = paste0(unique(dx_type), collapse = ';'),
            first_dx_name = paste0(unique(dx_name), collapse = ';'),
            .groups='drop') %>%
  left_join(pat_cohort %>% 
              select(pat_id, index_enc_date, final_censor_date, final_censor_type, ltfu_status, ltfu_reengage_date)
            , by="pat_id") %>%
  mutate(time_to_first_dx = difftime(first_dx_date, index_enc_date, units="days"))
```

```{r uf_first_dx_plot}
uf_first_dx %>%
  group_by(first_dx_date) %>%
  summarise(count=n(), .groups='drop') %>%
  mutate(cumulative_count = cumsum(count)) %>%
  ggplot(data=., aes(x=first_dx_date, y=cumulative_count)) +
  geom_line() +
  ylab('Cumulative Count of Patients Diagnosed with UF') +
  xlab('Date of Diagnosis') +
  theme_bw()
```

```{r}
uf_first_dx %>%
  mutate(medical_hx = case_when(grepl('Medical History', first_dx_type)~TRUE, TRUE~FALSE),
         encounter = case_when(grepl('Encounter Diagnosis', first_dx_type)~TRUE, TRUE~FALSE),
         problem_list = case_when(grepl('Problem List', first_dx_type)~TRUE, TRUE~FALSE),
         hospital_dx = case_when(grepl('Hospital Account Diagnosis', first_dx_type)~TRUE, TRUE~FALSE),
         referral_dx = case_when(grepl('Referral Diagnosis', first_dx_type)~TRUE, TRUE~FALSE)) %>%
CreateTableOne(data=., vars = c('referral_dx', 'encounter', 'medical_hx', 'problem_list', 'hospital_dx'))
```

The table above shows the relative frequency of each code type for the first ever UF Dx code drop. These sum to more than 100% because patients can get one code type dropped on the same day as another (i.e. a referral Dx and an encounter Dx on the same day).

```{r time_to_first_uf_dx_plot}
#summary(as.numeric(uf_first_dx$time_to_first_dx)/7)
uf_first_dx %>%
  # only for those diagnosed under 6 months
  ggplot(data=., aes(x=time_to_first_dx)) +
  geom_histogram(fill='lavender', color='black', binwidth=30) +
  xlab('Days between Index Encounter and First UF Dx') +
  theme_bw()
```

There are `r length(which(uf_first_dx$time_to_first_dx <= 180))` patients diagnosed in their first 6 months of follow-up.

How many patients are censored prior to their first UF diagnosis code?

`r length(which(uf_first_dx$first_dx_date > uf_first_dx$final_censor_date))` patients are censored prior to their first UF Dx code gets dropped.

How are these patients censored?

```{r}
uf_first_dx %>%
  filter(first_dx_date > final_censor_date) %>%
  CreateTableOne(data=., vars='final_censor_type')
```

The majority of them are lost to follow-up (LTFU). Of those LFTU, how many of them re-engage?

```{r}
tab <- uf_first_dx %>%
  filter(first_dx_date > final_censor_date & final_censor_type == 'LTFU') %>%
  mutate(ltfu_status = factor(ltfu_status)) %>%
  CreateTableOne(data=., vars='ltfu_status')

print(tab, showAllLevels=TRUE)
```

Very few of these patients are completely LTFU. If this was coded correctly, they should re-engage prior to their first UF diagnosis. Double checking this below...

```{r}
uf_first_dx %>%
  filter(first_dx_date > final_censor_date & final_censor_type == 'LTFU' & ltfu_status=='partial') %>%
  mutate(reengage_prior_to_dx = as.numeric(difftime(first_dx_date, ltfu_reengage_date, units='days'))) %>%
  ggplot(data=., aes(x=reengage_prior_to_dx)) +
  geom_histogram(fill='lightpink', color='black', binwidth=30) +
  xlab("Days between First UF Diagnosis date and Re-Engagement in relevant care") +
  theme_bw()
```

The graphs above shows that some patients received their first UF diagnosis sometime between the last known relevant care encounter prior to LTFU (3 years of no relevant care) and the next time they have relevant care (sometime greater than 3 years later). This would suggest these patients are still receiving SOME kind of care that allowed them to get a diagnosis code dropped but we do not consider this relevant care in our definition. 

```{r}
uf_first_dx %>%
  filter(first_dx_date > final_censor_date & final_censor_type == 'LTFU') %>%
  mutate(medical_hx = case_when(grepl('Medical History', first_dx_type)~TRUE, TRUE~FALSE),
         encounter = case_when(grepl('Encounter Diagnosis', first_dx_type)~TRUE, TRUE~FALSE),
         problem_list = case_when(grepl('Problem List', first_dx_type)~TRUE, TRUE~FALSE),
         hospital_dx = case_when(grepl('Hospital Account Diagnosis', first_dx_type)~TRUE, TRUE~FALSE),
         referral_dx = case_when(grepl('Referral Diagnosis', first_dx_type)~TRUE, TRUE~FALSE)) %>%
  CreateTableOne(data=., test=FALSE,
                 strata = 'ltfu_status', 
                 vars = c('referral_dx', 'encounter', 'medical_hx', 'problem_list', 'hospital_dx'))
```
Many of these diagnoses that occur after complete or partial diagnoses are encounter diagnoses. About one-fourth are referral diagnoses which is expected and another fourth to third are medical history diagnoses which could once again be indicative of being diagnosed outside of NYU and indicating this new diagnosis upon return to care at NYU or at 'irrelevant' care at NYU.

### Multiple diagnoses & count of referral diagnosis

```{r uf_multiple_dx}
uf_dx_count <- uf_all_long %>% 
  mutate(referral_dx=ifelse(grepl('Referral Diagnosis', event_type), TRUE, FALSE),
         medical_hx=ifelse(event_type == 'Medical History', TRUE, FALSE)) %>%
  group_by(pat_id) %>%
  summarise(dx_count = sum(grepl('Diagnosis|Problem|History', event_type)), 
            referral_count = sum(referral_dx),
            all_medical_hx = ifelse(n()==sum(medical_hx), TRUE, FALSE),
            .groups='drop')
```

```{r uf_dx_count_plot}
ggplot(data=uf_dx_count, aes(x=dx_count)) + 
  geom_histogram(fill='lightblue', color='black', binwidth = 5) +
  xlab('Number of UF Dx per patient') +
  theme_bw()
```

`r length(which(uf_dx_count$dx_count == 1))` cases have only 1 UF diagnosis code in their records.
This is `r round(length(which(uf_dx_count$dx_count == 1))/n_uf_raw, 2)*100`% of UF patients. Some of these may be confirmed below via other hints.

```{r}
uf_all_long %>%
  filter(pat_id %in% uf_dx_count$pat_id[which(uf_dx_count$dx_count == 1)]) %>%
  filter(grepl('Diagnosis|History|Problem', event_type)) %>%
  mutate(medical_hx = case_when(grepl('Medical History', event_type)~TRUE, TRUE~FALSE),
         encounter = case_when(grepl('Encounter Diagnosis', event_type)~TRUE, TRUE~FALSE),
         problem_list = case_when(grepl('Problem List', event_type)~TRUE, TRUE~FALSE),
         hospital_dx = case_when(grepl('Hospital Account Diagnosis', event_type)~TRUE, TRUE~FALSE),
         referral_dx = case_when(grepl('Referral Diagnosis', event_type)~TRUE, TRUE~FALSE)) %>%
  CreateTableOne(data=., test=FALSE,
                 vars = c('referral_dx', 'encounter', 'medical_hx', 'problem_list', 'hospital_dx'))
```
Of those with only 1 UF diagnosis code, most are encounter diagnoses and not medical history diagnoses which is promising. The 277 patients with only one diagnosis code, that is a medical history, are going to be difficult to place and may be prevalent cases. 

```{r}
tab <- uf_first_dx %>%
  filter(pat_id %in% uf_dx_count$pat_id[which(uf_dx_count$dx_count == 1)]) %>%
  filter(first_dx_date > final_censor_date & final_censor_type == 'LTFU') %>%
  mutate(ltfu_status = factor(ltfu_status)) %>%
  CreateTableOne(data=., vars='ltfu_status')

print(tab, showAllLevels=TRUE)
```
Most patients with only one UF dx who are also are diagnosed after LTFU are partially LTFU (they are likely these medical history diagnosis after re-engage patients).

```{r}
tab <- uf_first_dx %>%
  left_join(uf_dx_count, by='pat_id') %>%
  filter(first_dx_date > final_censor_date & ltfu_status == 'complete') %>%
  mutate(dx_count = factor(dx_count)) %>%
  CreateTableOne(data=., vars=c('dx_count', 'first_dx_type'), test=FALSE)

print(tab, showAllLevels=TRUE)
```
For complete LTFU patients diagnosed after censoring, most of them only have one diagnosis code. They are also predominantly encounter/referral diagnoses and hospital account diagnoses. Some are medical history diagnoses as well. 

### Only medical history codes (hint for prevalent case)

```{r medical_hx_only_hint}
# look for patients who ONLY ever have medical history code drops (no procedures)
uf_only_medical_hx <- uf_all_long %>% 
  group_by(pat_id) %>%
  summarise(all_medical_hx = ifelse(n()==sum(event_type == 'Medical History'), TRUE, FALSE),
            .groups='drop')
```

There are `r length(which(uf_only_medical_hx$all_medical_hx))` patients for whom the only indication for UF in their EHR is via medical history codes (and no diagnostic procedures). There is a high likelihood these patients are prevalent cases who only had their health history added to their record after entry to cohort. 

Assumption: if they were diagnosed outside NYU but during follow-up, they'd have a referral diagnosis code. 

```{r medical_hx_age_plot}
# are these patients generally older
pat_cohort %>%
  filter(pat_id %in% uf_only_medical_hx$pat_id[which(uf_only_medical_hx$all_medical_hx)]) %>%
  ggplot(data=., aes(x=age_20160801)) +
  geom_histogram(fill='lightyellow', color='black', binwidth=5) +
  xlab('Age at basline') +
  theme_bw()
```

Patients who only ever get UF codes dropped as medical history (and never have any procedures) are predominantly older patients (above 35 years old). I will explore the younger patients but I think in general these patients should be considered prevalent cases and exclude from the analysis. 

```{r, eval=FALSE}
# look at younger patients in this subgroup
uf_all_long %>% 
  filter(pat_id %in% uf_only_medical_hx$pat_id[which(uf_only_medical_hx$all_medical_hx)]) %>%
  # want to know if these are older patients
  left_join(pat_cohort, by='pat_id') %>%
  filter(age_20160801 < 35) %>%
  View()
```

These are a bit tricky to discern right now. The question is: is there anyway an incident diagnosis of fibroids (at NYU or outside of NYU) would only ever receive a medical history code in their EHR?

```{r}
# time to first dx for all medical history cases
uf_first_dx %>%
  filter(pat_id %in% unique(uf_only_medical_hx$pat_id[which(uf_only_medical_hx$all_medical_hx)])) %>%
  ggplot(data=., aes(x=as.numeric(time_to_first_dx))) +
  geom_histogram(fill='lavender', color='black', binwidth=30) +
  xlab('Days between Index Encounter and First UF Dx') +
  geom_vline(xintercept = 365, color='orange') +
  theme_bw()
```
Some of these patients are surely prevalent UF cases based on the early time to first diagnosis, but there is a wide distribution as to the timing of the first diagnoses which implies that some are incident cases that for some reason only get dropped as medical history diagnoses from the start. Assumption: these represent cases diagnosed outside of NYU and then come to NYU after treatment/diagnosis. 

### Diagnostic procedure hints

```{r, warnings=FALSE}
df <- uf_all_long
setDT(df)
# indexing which rows are procedures
df$is_procedure <- grepl('Imaging|Surgery',df$event_type)
# indexing which rows are dx (ignore imaging Dx for now)
df$is_diagnosis <- grepl('Diagnosis|Problem|List|History',df$event_type) & !grepl('Imaging',df$event_type)
# rearrange so if a procedure and dx happen on the same day the procedure goes first
df <- arrange(df, pat_id, date, is_diagnosis, is_procedure)
# get the index of the next row that is a dx when the row is a procedure
df[,next_dx_index:=zoo::na.locf(NA^(!is_diagnosis)*.I, fromLast=T, na.rm=F)*NA^(!is_procedure), by=pat_id]
# find time difference between the date of procedure and the next Dx date
df <- df[,next_dx_time_interval:=difftime(date[next_dx_index], date, units="days")]
df <- df[,next_dx_name:=event_name[next_dx_index]]
df <- df[,next_dx_type:=event_type[next_dx_index]]
# check this works
#View(df)

# which are on the same day
length(which(df$next_dx_time_interval==0))
```

```{r}
df2 <- as.data.frame(df) %>%
  mutate(method = case_when(grepl('Surgery', event_type)~"Surgery",
                            grepl('Imaging', event_type)~"Imaging",
                            TRUE~NA))

df2 %>%
  filter(!is.na(next_dx_time_interval)) %>%
  group_by(pat_id) %>%
  filter(next_dx_time_interval == min(next_dx_time_interval, na.rm=TRUE)) %>%
  # in the case that there are multiple procedures with same margin, take the earlier one
  filter(date == min(date, na.rm=TRUE)) %>%
  ungroup() %>%
  ggplot(data=., aes(x=next_dx_time_interval/30, fill=method)) +
  xlab('Shortest Time between Dx procedure and next Dx code (months)') +
  geom_histogram(binwidth = 5) +
  facet_wrap(~method, scales = 'free_y') +
  theme_bw()
```
The graph shows the count for patients who have an UF Dx, had any diagnostic procedure performed in their EHR, and had another UF Dx code following that procedure. The earliest procedure with the shortest margin between procedure and Dx code are selected per patient. The vast majority have an imaging procedure performed and within one month another Dx code is dropped. For those with a surgical procedure chosen, they are also quickly followed with another Dx code.

## Exploring different types of confident diagnosis

```{r}
confident_uf_dx <- uf_dx_unique %>%
  distinct(pat_id) %>%
  mutate(confident_dx = FALSE)
```

### They have any Dx code type dropped within 6 months of a diagnostic procedure

```{r}
# grab ids for those who have any procedure with another diagnosis code within
ids <- unique(df2$pat_id[which(df2$next_dx_time_interval <= 180)])
# change their status in dataframe and add type
confident_uf_dx <- confident_uf_dx %>%
  mutate(confident_dx = case_when(pat_id %in% ids~TRUE,
                                  TRUE~confident_dx),
         confirmed_via_procedure = case_when(pat_id %in% ids~TRUE,
                                  TRUE~FALSE))
```

### They have location of the fibroid

Assumption: the only way a location would be supplied is through some diagnostic procedure even if not noted in their charts. 

```{r}
# extract location of the UF based on Dx name if that exists
df2$location <- grepl('intramural|subserous|submucosal|subserosal|submucous|cervix|cervical', tolower(df2$event_name))

# grab patients who have any UF diagnosis with a location
# these could still be prevalent cases but we still assume these are true cases
ids <- unique(df2$pat_id[which(df2$location)])
```

```{r, eval=FALSE}
## location is a strong indicator of confirmed UF since we ASSUME location was ascertained through imaging at NYU or outside of NYU (referral)
df2 %>% filter(pat_id %in% ids) %>% View()

## look at whos leftover
df2 %>% 
  filter(!pat_id %in% ids & pat_id %in% confident_uf_dx$pat_id[which(!confident_uf_dx$confident_dx)]) %>%
  group_by(pat_id) %>% 
  filter(sum(location)>=1) %>%
  View()
```

```{r}
## add these to the confident_dx dataframe
confident_uf_dx <- confident_uf_dx %>%
  mutate(confident_dx = case_when(pat_id %in% ids~TRUE, 
                                  TRUE~confident_dx),
         location_of_fibroids = case_when(pat_id %in% ids~TRUE,
                                  TRUE~FALSE))
```

## They have fibroids added to medical history/problem list after any diagnostic procedure

We might assume that prior to surgery someone might have a clinical fibroid diagnosis (unspecified location or no imaging confirmed location), but then after surgery the fibroids are treated as a prevalent diagnosis

```{r, eval=FALSE, include=FALSE}
# example of patient with no confirmatory surgery (no Dx closely following surgery/imaging)
#uf_all_long %>% filter(pat_id == 'Z2405366') %>% View()

# they have multiple Dx codes (unspecific location -- so clinical)
# then they have imaging/surgery
# then years later it gets added to their problem list

# check out procedures followed by problem list/medical history drop
length(which(grepl("Problem|History", df$next_dx_type) & grepl("Surgery", df$event_type) & 
               df$next_dx_time_interval > 180))
ids <- unique(df$pat_id[which(grepl("Problem|History", df$event_type[df$next_dx_index]) & df$next_dx_time_interval > 180)])
#df2 %>% filter(pat_id %in% ids) %>% View()
```

```{r}
## add these to the confident_dx dataframe
confident_uf_dx <- confident_uf_dx %>%
  mutate(confident_dx = case_when(pat_id %in% ids~TRUE,
                                  TRUE~confident_dx),
         prevalent_dx_after_procedure = case_when(pat_id %in% ids~TRUE,
                                  TRUE~FALSE))

#summary(confident_uf_dx)
```

## They are diagnosed with uterine fibroids during pregnancy. 

Shilpi confirmed that most of these are likely incidental fibroid diagnoses resulting from imaging related to pregnancy. Therefore many of these cases may be asymptomatic but are true incidental cases nonetheless. Only count when they are not a 'medical history' diagnosis as this might refer to a prior pregnancy. 

```{r, eval=FALSE}
# how many of these are pregnancy diagnoses
df2 %>% 
  filter(pat_id %in% confident_uf_dx$pat_id[which(!confident_uf_dx$confident_dx)]) %>% 
  group_by(pat_id) %>%
  summarise(has_pregnancy_fibroids = ifelse(sum(grepl("pregnancy|partum|antenatal", tolower(event_name)))>=1, TRUE, FALSE),
            .groups='drop') %>%
  summary()

# look at diagnosis types for pregnancy diagnoses
df2 %>% 
  filter(grepl("pregnancy|partum|antenatal", tolower(event_name))) %>%
  mutate(medical_hx = case_when(grepl('Medical History', event_type)~TRUE, TRUE~FALSE),
         encounter = case_when(grepl('Encounter Diagnosis', event_type)~TRUE, TRUE~FALSE),
         problem_list = case_when(grepl('Problem List', event_type)~TRUE, TRUE~FALSE),
         hospital_dx = case_when(grepl('Hospital Account Diagnosis', event_type)~TRUE, TRUE~FALSE),
         referral_dx = case_when(grepl('Referral Diagnosis', event_type)~TRUE, TRUE~FALSE)) %>%
  CreateTableOne(data=., vars = c('referral_dx', 'encounter', 'medical_hx', 'problem_list', 'hospital_dx'))
```

```{r}
# grab ids for those with a pregnancy diagnosis (not medical history)
ids <- unique(df$pat_id[which(!grepl("History", df$event_type) & 
                                grepl("pregnancy|partum|antenatal|antepartum", tolower(df$event_name)))])

## add these to the confident_dx dataframe
confident_uf_dx <- confident_uf_dx %>%
  mutate(confident_dx = case_when(pat_id %in% ids~TRUE,
                                  TRUE~confident_dx),
         pregnancy_uf_dx = case_when(pat_id %in% ids~TRUE,
                                  TRUE~FALSE))

#summary(confident_uf_dx)
```

## Add multiple referral diagnosis codes as to confident dx

Assumption: if patients are being referred multiple times for UF -- this could indicate that they have strong symptomology and/or they are diagnosed by an external provider and then referred to NYU for treatment. 

```{r}
ids <- uf_dx_count$pat_id[which(uf_dx_count$referral_count > 1)]

## add these to the confident_dx dataframe
confident_uf_dx <- confident_uf_dx %>%
  mutate(confident_dx = case_when(pat_id %in% ids~TRUE,
                                  TRUE~confident_dx),
         multiple_referral_dx = case_when(pat_id %in% ids~TRUE,
                                  TRUE~FALSE))
```

## Add patients diagnosed after partial LTFU 

Assumption: Patients who receive their first UF Dx on the day they return to relevant care at NYU were likely diagnosed outside of NYU and return to NYU with the new incident diagnosis. 

QUESTION: Should patients diagnosed DURING partial LTFU be included here? Based on dicussion with Anais, these could be encounter diagnoses dropped for the referral (forced by Epic) but is not a confirmed diagnosis.

```{r}
# people diagnosed AFTER re-engaging in relevant care post-LTFU are likely incident cases
ids <- uf_first_dx$pat_id[which(uf_first_dx$first_dx_date==uf_first_dx$ltfu_reengage_date)]

# add to the confident_dx dataframe
confident_uf_dx <- confident_uf_dx %>%
  mutate(confident_dx = case_when(pat_id %in% ids~TRUE,
                                  TRUE~confident_dx),
         dx_after_partial_LTFU = case_when(pat_id %in% ids~TRUE,
                                  TRUE~FALSE))

#summary(confident_uf_dx)
```

## Add confident prevalent diagnoses based on Hx Dx codes at or close to first dx code

Assumption: If within two months of the first dx code they also get a code that indicates 'history of endometriosis' there is a good chance that the first diagnosis code was for a prevalent diagnosis. 

```{r}
## grab hx codes and record first instance
uf_hx_code <- uf_endo_cohort %>% 
  filter(grepl('Z86', icd10)) %>%
  group_by(pat_id) %>% 
  summarise(first_hx_code = min(dx_date), .groups='drop') %>%
  inner_join(uf_first_dx %>% select(pat_id, first_dx_date, first_dx_type, first_dx_name), by='pat_id') %>%
  mutate(hx_code_margin = difftime(first_hx_code, first_dx_date, unit='days'))

# are any before the index encounter date?
length(unique(uf_endo_cohort$pat_id[which(uf_endo_cohort$index_enc_date >= uf_endo_cohort$dx_date & 
                                            grepl('Z86', uf_endo_cohort$icd10))]))
# there are 0 -- good

## look at how many are close to the date of first dx code
prevalent_ids <- unique(uf_hx_code$pat_id[which(uf_hx_code$hx_code_margin<=60)])
ids <- unique(uf_hx_code$pat_id[which(uf_hx_code$hx_code_margin>60)])

# add to the confident_dx dataframe as true negatives
confident_uf_dx <- confident_uf_dx %>%
  mutate(confident_dx = case_when(pat_id %in% c(prevalent_ids, ids)~TRUE,
                                  TRUE~confident_dx),
         no_prevalent_hx_code = case_when(pat_id %in% ids~TRUE,
                                  TRUE~FALSE),
         confident_prevalent = case_when(pat_id %in% prevalent_ids~TRUE,
                                  TRUE~FALSE))
```

## Look at final confident dx table

```{r}
print(CreateTableOne(data=confident_uf_dx, includeNA=TRUE,
               vars=c('confident_dx', 'confirmed_via_procedure', 'location_of_fibroids',
                      'prevalent_dx_after_procedure','pregnancy_uf_dx', 'multiple_referral_dx',
                      'dx_after_partial_LTFU', 'no_prevalent_hx_code', 'confident_prevalent')),
      showAllLevels = TRUE)
```
## Look at patients who have no confident diagnosis based on above

```{r}
# grab IDs of patients who never get any procedure during follow-up
ids <- confident_uf_dx$pat_id[which(!confident_uf_dx$confident_dx)]

# grab characteristics from the procedure dataframe
procedure_info <- df2 %>%
  group_by(pat_id) %>%
  summarise(
    has_any_imaging=ifelse(sum(method=='Imaging', na.rm=TRUE)>=1, TRUE, FALSE),
    has_any_surgery=ifelse(sum(method=='Surgery', na.rm=TRUE)>=1, TRUE, FALSE),
    shortest_procedure_dx_margin = ifelse((all(is.na(next_dx_time_interval))), NA, 
                                          min(next_dx_time_interval, na.rm = TRUE)), 
    .groups='drop')
 
# join together this and other information on these patients 
df_not_confident <- uf_first_dx %>%
  left_join(uf_dx_count, by='pat_id') %>%
  left_join(procedure_info, by='pat_id') %>%
  filter(pat_id %in% ids) %>%
  mutate(first_dx_medical = grepl('Medical History', first_dx_type),
         only_one_code=ifelse(dx_count==1, TRUE, FALSE),
         time_to_first_dx=as.numeric(time_to_first_dx),
         shortest_procedure_dx_margin=as.numeric(shortest_procedure_dx_margin),) %>%
  mutate(dx_early_into_FU = ifelse(time_to_first_dx <= 180, TRUE, FALSE))

# look at overall characteristics
tab_not_confident <- CreateTableOne(data=df_not_confident, 
                                    vars=c("first_dx_medical", "dx_early_into_FU", "dx_count", "referral_count", 
                                           "all_medical_hx", "has_any_surgery", "has_any_imaging", 
                                           "shortest_procedure_dx_margin", "only_one_code"))

print(tab_not_confident, 
      nonnormal=c("dx_count", "referral_count", "shortest_procedure_dx_margin"), 
      showAllLevels = TRUE)
```

```{r not_confident_hints_corrplot}
# make correlation matrix for non-confident case characteristics
corrplot.mixed(cor(df_not_confident %>% 
                     select(dx_early_into_FU, only_one_code, referral_count, 
                            all_medical_hx, has_any_imaging, has_any_surgery), use='pairwise.complete.obs'),  
               lower = "number", 
               upper = "ellipse",
               tl.pos = "lt",
               tl.col = "black") 
```
Based on looking at this correlation matrix, the most relevant characteristics of the non-confident cases appears to be having only one code, number of referrals, and whether all dx codes are medical history. Therefore chart review will be broken down as follows:

- Only one code (n=80)
- More than one code & all medical hx codes (n=30) --> likely prevalent cases
- More than one code & NOT all medical hx codes (n=90)

Furthermore, given the relevance of referral codes, we will explore the assumption that 2 or more referral dx codes are incident cases (n=20). Finally, we will explore the assumption that first dx after partial LTFU are incident cases (n=20).

## Generate subset of non-confident cases for chart review

```{r}
set.seed(2024)
## grab random subset of cases from each group of non-confident cases
chart_review_cases <- df_not_confident %>% 
  filter(only_one_code) %>% sample_n(., 80) %>% 
  mutate(group = 'only_one_code') %>%
  bind_rows(df_not_confident %>% filter(!only_one_code) %>% sample_n(., 90) %>% 
              mutate(group = 'more_than_one_code')) %>%
  bind_rows(df_not_confident %>% filter(all_medical_hx) %>% sample_n(., 30) %>% 
              mutate(group = 'all_medical_hx')) %>%
  bind_rows(confident_uf_dx %>% filter(multiple_referral_dx) %>% sample_n(., 20) %>% 
              mutate(group = 'multiple_referrals')) %>%
  bind_rows(confident_uf_dx %>% filter(dx_after_partial_LTFU) %>% sample_n(., 20) %>% 
              mutate(group = 'dx_after_partial_LTFU')) %>%
  left_join(pat_cohort %>% select(pat_id, pat_mrn_id), by='pat_id') %>%
  select(pat_mrn_id, pat_id:group)

## write out file
write.csv(chart_review_cases, 'Data/Derived/uf_chart_review_cases.csv', row.names = FALSE)
```


## Look at non-confident diagnoses

```{r}
# look at those who are not confident Dx
df2 %>% 
  filter(pat_id %in% confident_uf_dx$pat_id[which(!confident_uf_dx$confident_dx)]) %>% 
  View()
```

```{r}
# how many of these are only medical history diagnoses
confident_uf_dx %>% 
  left_join(uf_only_medical_hx, by='pat_id') %>%
  filter(!confident_dx) %>%
  summary()
```

```{r}
# make table of distribution
tab_margin <- df2 %>%
  rename(event_margin = next_dx_time_interval) %>%
  filter(!is.na(event_margin)) %>%
  group_by(pat_id) %>%
  filter(event_margin == min(event_margin, na.rm=TRUE)) %>%
  # in the case that there are multiple procedures with same margin, take the earlier one
  filter(date == min(date, na.rm=TRUE)) %>% #filter(n()>1)
  ungroup() %>%
  mutate(event_margin_cat = case_when(event_margin == 0 ~ "Same day",
                                      event_margin <= 31 ~ "Same month",
                                      event_margin <= 90 ~ "Within 3 months",
                                      event_margin <= 180 ~ "Within 6 months",
                                      event_margin <= 365 ~ "Within 1 year",
                                      event_margin <= 730 ~ "Within 2 years",
                                      event_margin > 730 ~ "More than 2 years")) %>%
  mutate(event_margin_cat = factor(event_margin_cat, levels = c('Same day', 'Same month', 
                                                                'Within 3 months', 'Within 6 months',
                                                                'Within 1 year', 'Within 2 years',
                                                                'More than 2 years'))) %>%
  CreateTableOne(vars="event_margin_cat", strata="method", data=., test=FALSE)

# print table
print(tab_margin, nonnormal="event_margin")
```

## Look at patients who have no procedures

```{r}
# grab IDs of patients who never get any procedure during follow-up
ids <- df2 %>% group_by(pat_id) %>% 
  filter(sum(is_procedure)==0) %>% 
  select(pat_id)
ids <- unique(ids$pat_id)
  
# look at overall characteristics
tab_no_procedure <- uf_first_dx %>%
  left_join(uf_dx_count, by='pat_id') %>%
  filter(pat_id %in% ids) %>%
  mutate(time_to_first_dx=as.numeric(time_to_first_dx)) %>%
  CreateTableOne(data=., vars=c("time_to_first_dx", "dx_count", "referral_count", "all_medical_hx"))

print(tab_no_procedure, nonnormal=c("time_to_first_dx", "dx_count", "referral_count"), showAllLevels = TRUE)

# look at all diagnoses for these patients
#df2 %>% group_by(pat_id) %>% filter(sum(is_procedure)==0) %>% View()
```

## Look at time to first dx for all medical hx cases

```{r}
summary(df_not_confident$time_to_first_dx[which(df_not_confident$all_medical_hx)])
```

It doesn't appear that patients who only receive medical history diagnoses are early diagnosed cases in the cohort. The next step is look into whether these remaining not confident cases are censored and therefore, likely to have recieved an external diagnosis and treatment, explaining the paucity of their diagnosis information in the NYU EHR. 

## Look at censoring date and first diagnosis date for these non-confident cases

```{r}
df_not_confident %>%
  mutate(censored_before_dx = ifelse(first_dx_date >= final_censor_date, TRUE, FALSE)) %>%
  CreateTableOne(data=., var=c('final_censor_type', 'censored_before_dx'), includeNA=TRUE)
```
About 60% of non-confident cases experience some form of censoring and about 20% of these cases are diagnosed after censoring occurred. This could likely indicate more usage of other healthcare systems in which an incident UF dx occurred and then was at some point communicated back to NYU. 

```{r}
# look at those diagnosed after LTFU censoring
```

Those who age out or experience menopause might be more likely to be prevalent cases -- check if they are fall into the all medical history diagnosis category. 

```{r}
# look at censoring among those with all medical history codes
df_not_confident %>%
  filter(all_medical_hx) %>%
  mutate(censored_before_dx = ifelse(first_dx_date >= final_censor_date, TRUE, FALSE)) %>%
  CreateTableOne(data=., var=c('final_censor_type', 'censored_before_dx'), includeNA=TRUE)
```
Surprisingly, the majority of the non-confident cases with only medical history diagnoses are not predominantly aged out or menopause censored. In fact, many are never censored. 

```{r}
# look at them
df_not_confident %>% filter(all_medical_hx==TRUE) %>% View()
```

From reviewing the all medical history non-confident cases, it appears to me that many of them are likely patients who predominantly receive their UF diagnosis and treatment outside NYU. About 1/4 of these patients 

Finally, those who are censored via surgery may be true cases referred in for surgery with little diagnosis information in NYU EHR -- also look into these. 

```{r}
# look into those censored due to surgery
```

```{r}
# subset to only those with a UF dx
uf_dx_hints <- uf_hints %>% 
  filter(uf_dx) %>% 
  select(pat_id, first_dx_referral:surgery_w_dx_after) %>%
  mutate(number_hints = rowSums(.[2:10], na.rm=TRUE))
```

```{r uf_hints_corrplot}
corrplot.mixed(cor(uf_dx_hints[,-1], use='pairwise.complete.obs'),  # Correlation matrix
               lower = "number", 
               upper = "ellipse",
               tl.pos = "lt",
               tl.col = "black") 
```
The correlation plot above visualizes the correlation between these binary UF diagnosis characteristics. For example, the presence of imaging is negatively correlated (-0.47) with the presence of surgery in someone's EHR. My interpretation here is that many patients are referred into NYU for surgery for UF but these patients are therefore much less likely to get imaging. Other important observations include:

- confirmed imaging has a positive correlation with surgery and confirmation via surgery (these are highly confident cases)
- confirmation via surgery is negatively correlated with later Dx codes after 3 months of surgery (this makes sense if we think about surgery as treatment for the condition, although recurrence is possible)
- the first dx code being a referral Dx is positively correlated with having multiple Dx codes and other referral Dx codes
- confirmation via imaging OR surgery is positively correlated with have multiple Dx codes and multiple referral Dx codes
- the hint with the weakest relationship to the overall number of hints a patient has is the presence of imaging (this makes sense since people could get imaging for any number of reasons)

```{r}
uf_dx_hints %>%
  ggplot(data=., aes(x=number_hints)) +
  geom_histogram(fill='darkseagreen2', color='black', binwidth=1) +
  xlab('Number of UF Dx Hints per patient') +
  theme_bw()
```
