---
title: "Step 2.03: Symptom Outcome Exploration"
author: "Mia Charifson"
date: "2024-02-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/miacharifson/Library/CloudStorage/OneDrive-NYULangoneHealth/Charifson Dissertation/')

library(tidyverse)
library(ggplot2)
library(lubridate)
library(tableone)
library(arsenal)
library(readxl)
library(ggpubr)
library(corrplot)
library(data.table)
library(forcats)
```

## Introduction

The goal of this document is to: 

1. define the timing of the first symptom related to UF and Endo symptomology for each patient (never, before entry, at baseline, during follow-up)
2. explore the UF and Endo symptomolgy in our cohort population (which symptom, how many times)

```{r load}
pat_cohort <- read.csv('Data/Derived/pat_cohort_w_censoring.csv')
menstrual_symptoms <- read.csv('Data/Epic/table3_dx_menstrual_symptoms.csv')
other_symptoms <- read.csv('Data/Epic/table3_dx_other_symptoms.csv')
```

```{r join_symptoms}
symptoms <- bind_rows(menstrual_symptoms, other_symptoms) %>%
  inner_join(pat_cohort %>% select(pat_id, index_enc_date), by='pat_id')
```

## First, some cleaning

There are `r length(which(symptoms$dx_date == 'NULL'))` entries where the date of the diagnosis code is missing. All of these are `Medical History` type diagnoses. These are generally uninformative without a diagnosis date. 

TO DO: try to recover diagnosis dates for as many of these rows as possible, by joining on encounter ID.

```{r, eval=FALSE}
table(symptoms$dx_type[which(symptoms$dx_date=='NULL')])

symptoms %>%
  group_by(pat_id) %>%
  filter(n() == sum(dx_type == 'Medical History')) %>%
  arrange(pat_id)
```

## When do patients have their first UF or Endo related symptom?

There are `r length(unique(symptoms$pat_id))` patients in our cohort who have at least one UF or Endo related symptom in their EHR.

```{r first_symptom}
first_symptom <- symptoms %>%
  group_by(pat_id) %>%
  summarise(index_enc_date = min(index_enc_date),
            dx_date = min(dx_date, na.rm=TRUE),
            dx_names = paste0(unique(dx_name),collapse = ', '),
            icd10s = paste0(unique(icd10),collapse = ', '),
            n_specific_symptoms = sum(!grepl('M54.5|R19.7', icd10)),
            .groups='drop') %>%
  mutate(dx_date = as.Date(dx_date), 
         index_enc_date = as.Date(index_enc_date), 
         time_group = case_when(dx_date<index_enc_date~'before baseline', 
                                dx_date>=index_enc_date & dx_date<=(ymd(index_enc_date) + years(1))~'at baseline',
                                dx_date>(ymd(index_enc_date)+years(1))~'during follow-up',
                                TRUE~'has symptom, no date'),
         flag = case_when(grepl('M54.5|R19.7', icd10s) & n_specific_symptoms == 0~TRUE,
                          TRUE~FALSE))

flag_ids <- unique(first_symptom$pat_id[which(first_symptom$flag)])
```

```{r first_symptom_table, results='asis'}
pat_cohort %>%
  select(pat_id) %>%
  left_join(first_symptom %>% filter(!flag), by='pat_id') %>%
  mutate(time_group = ifelse(is.na(time_group), 'never', time_group)) %>%
  tableby(~time_group, data=.) %>%
  summary()
```

## For patients who have symptoms, how many symptoms do they report since their first one?

A distinct symptom is counted as a unique diagnosis date-ICD10 code combinations.

```{r symptom_count_fig}
symptoms %>%
  filter(!pat_id %in% flag_ids) %>%
  group_by(pat_id) %>%
  summarise(count_symptoms = n_distinct(dx_date, icd10), .groups='drop') %>%
  ggplot(aes(x=count_symptoms), data=.) +
  geom_histogram(fill= 'gray', color='black', bins = 100) +
  theme_bw()
```

```{r symptom_count_table, results='asis'}
symptoms %>%
  filter(!pat_id %in% flag_ids) %>%
  group_by(pat_id) %>%
  summarise(count_symptoms = n_distinct(dx_date, icd10), .groups='drop') %>%
  tableby(~count_symptoms, data=., numeric.stats = c('mean', 'median', 'q1q3', 'range')) %>%
  summary()
```

## Which symptoms are commonly reported in this cohort?

```{r}
symptom_wide_boolean <- symptoms %>%
  filter(!pat_id %in% flag_ids) %>%
  group_by(pat_id, dx_date) %>%
  summarise(Pelvic_pain = (sum(grepl('R10.2', icd10))>=1),
         Dysmenorrhea = (sum(grepl('N94.[4-6]', icd10))>=1),
         Dyspareunia = (sum(grepl('F52.6|N94.1', icd10))>=1),
         Infertility = (sum(grepl('N97.9', icd10))>=1),
         Intermenstrual_pain = (sum(grepl('N94.0', icd10))>=1),
         Ovarian_cysts = (sum(grepl('N83.0|N83.2', icd10))>=1),
         Anemia = (sum(grepl('D50.0', icd10))>=1),
         Back_pain = (sum(grepl('M54.5', icd10))>=1),
         Postcoital_bleed = (sum(grepl('N93.0', icd10))>=1),
         AUB = (sum(grepl('N93.8|N93.9', icd10))>=1),
         Diarrhea = (sum(grepl('R19.7', icd10))>=1),
         Menorrhagia = (sum(grepl('N92.0|N92.1', icd10))>=1),
         Bladder_rectum_symptoms = (sum(grepl('N32.9|K59.00', icd10))>=1),
         Urinary_urgency = (sum(grepl('N39.3', icd10))>=1),
         .groups='drop')
```

```{r}
patient_counts <- symptom_wide_boolean %>%
  summarise(Pelvic_pain = n_distinct(pat_id[Pelvic_pain]),
         Dysmenorrhea = n_distinct(pat_id[Dysmenorrhea]),
         Dyspareunia = n_distinct(pat_id[Dyspareunia]),
         Infertility = n_distinct(pat_id[Infertility]),
         Intermenstrual_pain = n_distinct(pat_id[Intermenstrual_pain]),
         Ovarian_cysts = n_distinct(pat_id[Ovarian_cysts]),
         Anemia = n_distinct(pat_id[Anemia]),
         Back_pain = n_distinct(pat_id[Back_pain]),
         Postcoital_bleed = n_distinct(pat_id[Postcoital_bleed]),
         AUB = n_distinct(pat_id[AUB]),
         Diarrhea = n_distinct(pat_id[Diarrhea]),
         Menorrhagia = n_distinct(pat_id[Menorrhagia]),
         Bladder_rectum_symptoms = n_distinct(pat_id[Bladder_rectum_symptoms]),
         Urinary_urgency = n_distinct(pat_id[Urinary_urgency])) %>%
  pivot_longer(cols=everything(), names_to='Symptom', values_to='Count') %>%
  mutate(Percent = paste0(round(Count/nrow(pat_cohort)*100, 2), '%'))

patient_counts %>%
  ggplot(aes(x=Count, y=fct_reorder(Symptom, Count), fill=Symptom), data=.) +
  geom_col() + 
  geom_label(aes(label=Percent), fill='white', label.padding = unit(0.1, "lines")) +
  theme_bw() + 
  ylab('Symptom Name') +
  xlab('Count of unique patients') +
  theme(legend.position = 'none')
```

```{r}
symptom_wide_boolean %>% 
  summarise(Pelvic_pain = sum(Pelvic_pain),
         Dysmenorrhea = sum(Dysmenorrhea),
         Dyspareunia = sum(Dyspareunia),
         Infertility = sum(Infertility),
         Intermenstrual_pain = sum(Intermenstrual_pain),
         Ovarian_cysts = sum(Ovarian_cysts),
         Anemia = sum(Anemia),
         Back_pain = sum(Back_pain),
         Postcoital_bleed = sum(Postcoital_bleed),
         AUB = sum(AUB),
         Diarrhea = sum(Diarrhea),
         Menorrhagia = sum(Menorrhagia),
         Bladder_rectum_symptoms = sum(Bladder_rectum_symptoms),
         Urinary_urgency = sum(Urinary_urgency)) %>%
  pivot_longer(cols=everything(), names_to='Symptom', values_to='Count') %>%
  ggplot(aes(x=Count, y=fct_reorder(Symptom, Count), fill=Symptom), data=.) +
  geom_col() +
  theme_bw() + 
  ylab('Symptom Name') +
  xlab('Count of unique instances') +
  theme(legend.position = 'none')
```

## Which symptom is commonly reported for the first symptom at baseline & during follow-up?

```{r}
n_timegroup <- table(first_symptom$time_group)
patient_counts <- first_symptom %>%
  filter(!flag) %>%
  group_by(time_group) %>%
  summarise(Pelvic_pain = (sum(grepl('R10.2', icd10s))),
         Dysmenorrhea = (sum(grepl('N94.[4-6]', icd10s))),
         Dyspareunia = (sum(grepl('F52.6|N94.1', icd10s))),
         Infertility = (sum(grepl('N97.9', icd10s))),
         Intermenstrual_pain = (sum(grepl('N94.0', icd10s))),
         Ovarian_cysts = (sum(grepl('N83.0|N83.2', icd10s))),
         Anemia = (sum(grepl('D50.0', icd10s))),
         Back_pain = (sum(grepl('M54.5', icd10s))),
         Postcoital_bleed = (sum(grepl('N93.0', icd10s))),
         AUB = (sum(grepl('N93.8|N93.9', icd10s))),
         Diarrhea = (sum(grepl('R19.7', icd10s))),
         Menorrhagia = (sum(grepl('N92.0|N92.1', icd10s))),
         Bladder_rectum_symptoms = (sum(grepl('N32.9|K59.00', icd10s))),
         Urinary_urgency = (sum(grepl('N39.3', icd10s)))) %>%
  pivot_longer(cols=c(Pelvic_pain:Urinary_urgency), names_to='Symptom', values_to='Count') %>%
  mutate(Percent = case_when(time_group == 'at baseline' ~ Count/n_timegroup[1],
                             time_group == 'during follow-up' ~ Count/n_timegroup[3],
                             time_group == 'before baseline' ~ Count/n_timegroup[2]), 
         Percent = paste0(round(Percent*100, 2), '%'))

patient_counts %>%
  filter(time_group %in% c('at baseline', 'during follow-up')) %>%
  ggplot(aes(x=Count, y=fct_reorder(Symptom, Count), fill=Symptom), data=.) +
  geom_col() + 
  geom_label(aes(label=Percent), fill='white', label.padding = unit(0.1, "lines")) +
  theme_bw() + 
  ylab('Symptom Name') +
  xlab('Count of unique patients') +
  theme(legend.position = 'none') +
  facet_wrap(~time_group)
```

## What symptoms are associated with each other?

```{r}
symptom_wide <- symptoms %>%
  filter(!pat_id %in% flag_ids) %>%
  group_by(pat_id, dx_date) %>%
  summarise(Pelvic_pain = (sum(grepl('R10.2', icd10))>=1),
         Dysmenorrhea = (sum(grepl('N94.[4-6]', icd10))>=1),
         Dyspareunia = (sum(grepl('F52.6|N94.1', icd10))>=1),
         Infertility = (sum(grepl('N97.9', icd10))>=1),
         Intermenstrual_pain = (sum(grepl('N94.0', icd10))>=1),
         Ovarian_cysts = (sum(grepl('N83.0|N83.2', icd10))>=1),
         Anemia = (sum(grepl('D50.0', icd10))>=1),
         Back_pain = (sum(grepl('M54.5', icd10))>=1),
         Post_coital_bleed = (sum(grepl('N93.0', icd10))>=1),
         AUB = (sum(grepl('N93.8|N93.9', icd10))>=1),
         Diarrhea = (sum(grepl('R19.7', icd10))>=1),
         Menorrhagia = (sum(grepl('N92.0|N92.1', icd10))>=1),
         Bladder_rectum_symptoms = (sum(grepl('N32.9|K59.00', icd10))>=1),
         Urinary_urgency = (sum(grepl('N39.3', icd10))>=1),
         .groups='drop')
```

```{r}
symptom_wide_count <- symptoms %>%
  filter(!pat_id %in% flag_ids) %>%
  group_by(pat_id) %>%
  summarise(Pelvic_pain = (sum(grepl('R10.2', icd10))),
         Dysmenorrhea = (sum(grepl('N94.[4-6]', icd10))),
         Dyspareunia = (sum(grepl('F52.6|N94.1', icd10))),
         Infertility = (sum(grepl('N97.9', icd10))),
         Intermenstrual_pain = (sum(grepl('N94.0', icd10))),
         Ovarian_cysts = (sum(grepl('N83.0|N83.2', icd10))),
         Anemia = (sum(grepl('D50.0', icd10))),
         Back_pain = (sum(grepl('M54.5', icd10))),
         Post_coital_bleed = (sum(grepl('N93.0', icd10))),
         AUB = (sum(grepl('N93.8|N93.9', icd10))),
         Diarrhea = (sum(grepl('R19.7', icd10))),
         Menorrhagia = (sum(grepl('N92.0|N92.1', icd10))),
         Bladder_rectum_symptoms = (sum(grepl('N32.9|K59.00', icd10))),
         Urinary_urgency = (sum(grepl('N39.3', icd10))),
         .groups='drop')
```

```{r}
corrplot.mixed(cor(symptom_wide_count[, -c(1)]), upper = 'square')
```

```{r}
require(graphics); require(grDevices)
x <- cor(symptom_wide_count[, -c(1)])
rc <- rainbow(nrow(x), start = 0, end = .3)
cc <- rainbow(ncol(x), start = 0, end = .3)
heatmap(x, RowSideColors = rc, ColSideColors = cc)
```

